Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-25 2024/OPTIMA/OPTIMA_2024_MCO.tex
fig added
Optima part 1

%
\begin{abstract}
To solve a multicriterial optimization problem, it is necessary to find a whole set of parameters values corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case of the criteria are multiextremal. In the paper discusses several varia-tions of the multicriteria optimization algorithm using the black box method with multiextremal criteria. The developed algorithms are based on an information-statistical approach and a machine learning procedure used to increase the efficiency of constructing the Pareto set. As a novelty, the authors proposed a new approach to using the constructed machine learning models. The method involves using the calculated probability of a trial points belonging to the Pareto set to assess the prospects one. The effectiveness of different variations of the developed algorithms estimated in a representative computational experiment.

@@

@@

@@

@@

@@

>>>>

%
\begin{abstract}
To solve a multicriterial optimization problem, it is necessary to find a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for solving black-box multiobjective problems with multiextremal criteria. The developed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriterial optimization. The effectiveness of different versions of the developed algorithm is estimated on the base of  a representative computational experiment.

@@

Problems of optimal choice are widespread in many areas of activity. As a rule, in complicated decision making models there are several functions to be optimized which reflect the decision making objectives and those are, in general case, contradictory. In these circumstances, the notion of optimal solution is more complicated than in single criterion (scalar) statement formulated as mathematical programming problem because there exist incomparable decisions. In this situation under introduced definition of domination, the set of non-dominated variants (Pareto set) is taken as the solution of the multiobjective problem.

In our study we deal with methods for solving the most complex class of multicriterial problems where criteria are considered as ``black-box'' functions and are multiextremal \cite{Miettinen1999,Ehrgott2005,Pardalos2017,Strongin2000,Sergeyev2013}.

It should be noted that problems of multiextremal optimization are among the most complicated problems of mathematical programming. Unlike the local optimization where an optimal point has to be juxtaposed with other points only in its region of attraction, the global solution must be better w.r.t. all points in the feasible domain. If to solve such problems by grid search methods, the amount of computations will rise exponentially when increasing the number of parameters (variables of the problem).

For solving the multiextremal problems the more effective algorithms are used which can be divided into two key classes, namely, into deterministic and meta-heuristic (nature-inspired) methods \cite{ML_MCO_2023,Evtushenko2014,Deb2002,Durillo2010,Mostaghim2007,NDG09,RC05,ZLT01,Gergel2018,GergelKozinov2020}.

Meta-heuristic algorithms model the behavior of living organisms or nature processes \cite{Deb2002,Durillo2010,Mostaghim2007,NDG09,RC05,ZLT01}. As a rule, these methods are simple enough in implementation and can demonstrate good efficiency for unimodal objective functions and multiextremal problems with a small number of local optima; however, they do not ensure finding the best solution.

Deterministic algorithms guarantee convergence to global optimum, but their computational schemes are more complex. However, advanced deterministic algorithms outperform metaheuristic ones in terms of quality \cite{Kvasov2018,Sergeyev2018}.

One of the qualitative deterministic methods for solving multiextremal problems is the  information-statistical algorithm of global search \cite{Strongin2000,Sergeyev2013}. This algorithm proposed initially for solving scalar problems was successfully developed for analyzing multiobjective statements \cite{ML_MCO_2023,Gergel2018,GergelKozinov2020}.

In the publication \cite{ML_MCO_2023} we proposed an approach to solving multicriterial optimization problems combining the classic information-statistical global search algorithm with machine learning methods. In the framework of the approach a problem of multicriterial optimization was transformed to a series of scalar optimization subproblems. In the course of solving scalar subproblems, on the base of machine learning methods a hyperplane was built and gradually refined which separated points of effective and ineffective solutions in the criteria space. Additional information connected with the separating hyperplane was embedded in the computational scheme of the optimization algorithm and, as a result, the number of trials (criteria evaluations) in domains with ineffective solutions was significantly decreased. The experimental results showed that in comparison with a number of heuristic algorithms, the search for effective solutions was several times faster. At the same time, this approach can be improved because utilizing only the separating plane limits the class of machine learning techniques which can be applied for enhancing the global search algorithm.

In the framework of our study we propose a new approach to separating search trial points in the criteria space based on the probability that reflects the belonging of the trial point to the Pareto set. The most machine learning models allow obtaining this metric which makes it possible to choose an algorithm for building a machine learning model depending on the features of the problem being solved.

The paper has the following structure. Section \ref{sec:2} describes the mathematical statement of the problem to be studied. Section \ref{sec:3} is devoted to a brief description of the basic algorithms used for solving problems of multicriterial optimization. Section \ref{sec:4} contains information on the techniques applied for reducing the number of trials during the optimization process. Subsection \ref{ssec:41} gives a short description of the method considered in \cite{ML_MCO_2023}. A new approach to using probabilities of trial points belonging to different efficiency classes is presented in subsection \ref{ssec:42}. Section \ref{sec:5} contains results of computational experiments demonstrating the effectiveness of the proposed approach.

@@

\subsection{Section 41}
\label{ssec:41}

\subsection{Section 42}
\label{ssec:42}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig1.png}
\label{fig1}
\caption{A figure caption.} 
\end{figure}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig2.png}
\label{fig2}
\caption{A figure caption.} 
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{fig3.png}
\label{fig3}
\caption{A figure caption.} 
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{fig4.png}
\label{fig4}
\caption{A figure caption.} 
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{fig5.png}
\label{fig5}
\caption{A figure caption.} 
\end{figure}
